\Name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Method for a Precision Spectrograph

### 2020-10-24 paper/paper.tex
Submitted Referee Report

\begin{abstract}\noindent%
\Name\ is a non-parametric, hierarchical framework for precision wavelength-calibration of spectrographs.  It is designed with the needs of extreme-precision radial velocity (\eprv) in mind, which require that instruments be calibrated or stabilized to better than $10^{-4}$ pixels.  Instruments vary along only a few dominant degrees of freedom, especially \eprv\ instruments that feature highly stabilized optical systems and detectors.  \Name\ takes advantage of this property by using all calibration data to construct a low-dimensional representation of all accessible calibration states for an instrument.  \Name\ also takes advantage of laser frequency combs or etalons, which generate a dense set of stable calibration points.  This density permits the use of a non-parametric wavelength solution that can adapt to any instrument or detector oddities better than parametric models, such as a polynomial.  We demonstrate the success of this method with data from the \textsl{EXtreme PREcision Spectrograph} (\expres), which uses a laser frequency comb.  When wavelengths are assigned to laser comb lines using \name, the RMS of the residuals is about five times lower than wavelengths assigned using polynomial fits to individual exposures.  Radial-velocity measurements of HD~34411 showed a reduction in RMS scatter over a 10-month time baseline from $1.22$ to $1.05\, \mps$.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- instrumentation: detectors -- techniques: spectroscopic -- techniques: radial velocities -- methods: data analysis -- methods: statistical}
%\section*{~}\pagebreak
\section{Introduction} 
Precise, radial-velocity programs have been fruitful in finding and characterizing extra-solar planets \citep[e.g.][]{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $50,000-100,000$, which correspond to line widths on the order of $3000\,\mps$.  The state of the art RV precision reached $1\, \mps$ in 2016 \added{\citep{fischer2016}}.  The newest generation of instruments aim  to reach $0.1\,\mps$ precision, the required precision to detect terrestrial worlds.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).  Two next-generation spectrographs, \expres\ and \espresso, have been commissioned for more than a year and are demonstrating $<0.1 \mps$ instrumental errors and $\sim 0.2\, \mps$ errors on stars \citep{pepe2013, jurgenson2016, blackman2020, petersburg2020, brewer2020, mascareno2020}.

@@

The second simple idea follows from the observation that most physical systems have only a few dominant degrees of freedom, meaning most spectrographs vary along only a small number of axes in ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  This is particularly true of \eprv\ instruments, which are equipped with stringent environmental stabilizing.  \replaced{ likely reducing the number and extent of the axes along which the instrument can vary.}{The thermomechanical stability of these instruments reduces the variations they experience to something that can be represented by a low-dimensional framework.}  That is, spectrographs, especially stabilized ones, should have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the calibration space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the small, accessible part of calibration space the spectrograph was located for each exposure.

@@

The challenge is to learn these basis functions, $g_k(\lambda,m)$, from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern the basis functions.  In this paper, we present one implementation of \name\ using principal component analysis (PCA) \added{\citep{pearson1901, jolliffe2016}}.  A PCA is justifiable in the limit where exposures have very high signal-to-nose ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

@@

\replaced{That is, \name\ assumes that this has been done correctly.  We caution, though, that systematic errors or large uncertainties in fitting line positions can propagate through to errors or biases in the wavelength models returned by \name.}{\Name\ assumes that line positions have been identified ``correctly,'' as in the position of a calibration line is determined the same way as the position of a stellar line when extracting RVs.  This also inherently assumes that the calibration lines are not subject to an effect that the science exposures are not, for example differences in charge transfer inefficiency, non-linearities in the PSF, etc.  We caution that systemic errors or large uncertainties in fitting line positions easily propagate through to biases in the wavelength models returned by \name.  For more discussion, see Section \ref{sec:discussion}.}

@@

\begin{algorithm}
\SetAlgoLined
\KwData{line positions $x(\lambda,m,n)$ for each exposure $n$, with wavelengths $\lambda$ and echelle orders $m$}
\KwResult{Basis vectors of the low-dimensional calibration space $g_k(\lambda,m)$ and location of exposures in calibration space expressed by amplitudes $a_{n,k}$}
\While{change in missing or outlier line centers $>$ 0.01\%}
{
	$g_0(\lambda,m) = \overline{x(\lambda,m,n)}$\;
	\added{using Singular-Value Decomposition,} find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x(\lambda,m,n)-g_0(\lambda,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(\lambda,m) = V$\;
	$x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)$ for $x(\lambda,m,n)$ = \code{NaN} where $K$ is a a small integer
	}
\caption{Dimensionality Reduction and De-Noising}
\end{algorithm}

@@

In the implementation of \name\ presented here, the amplitudes of the principal components are interpolated linearly with respect to time.  This is discussed more in Section \ref{sec:choice_avt}.  \added{Let a prime denote values related to a science exposure $n'$ for which we want wavelengths.}  We use linearly interpolated magnitudes, $a_{n',k}$ at time $t_{n'}$ to construct the calibration state of the spectrograph for that point in time.  Using interpolated amplitudes, $a_{n',k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, a new set of calibration lines, $x'(\lambda,m,n')$ can be constructed for any exposure as formulated in Equation \ref{eq:excl_wsol}.

@@

After experiments, we found that a cubic-spline interpolation that enforces monotonicity, such as a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a parameterized function, while the enforced monotonicity allows the wavelength solution, $\lambda(x',m,n')$, to be invertible and prevents spurious deviations that may befall a cubic spline.  Choices in interpolation scheme, $K$, and other tests are further discussed in Section \ref{sec:choice_wvp}.  \added{Our implementation of \name\ is hosted on GitHub\footnote{\url{https://www.github.com/davidwhogg/EPRVCalibration}}.}

\begin{algorithm}
\SetAlgoLined
\KwData{the fiducial calibration of the spectrograph $g_0(\lambda,m)$; magnitudes of the principal components for each exposure $a_{n,k}$; basis vectors spanning the calibration space of the spectrograph $g_k(\lambda,m)$; }
\KwResult{Wavelengths for detector positions $x'(m,n')$ of exposure $n'$ with time $t_{n'}$, \added{where the prime is used to denote values relevant to this new exposure}}

@@

@@

In order to \added{comfortably} satisfy the assumption that there exists only low-order variation, which is needed for \name, \replaced{we used exposures from after the photonic crystal fiber within the LFC was changed in Summer 2019, and the LFC stabilized}{we used exposures from after the LFC stabilized following servicing in summer 2019, where the photonic crystal fiber was replaced and the polarization was changed to shift the wavelength range of the LFC redwards}.  In the results presented here, we use 1227 LFC exposures and 78 ThAr exposures taken between October 14 and December 18, 2019 on 29 unique nights.

@@

\added{The polynomial and interpolation tests remove the same 50\% of lines from each exposure while the denoising test completely removes a randomly selected 10\% of calibration exposures and their associated line position measurements.  Errors from interpolation will be localized, extending only to neighboring lines.  We therefore aggressively remove every other line to ensure we are capturing these local effects.  The PCA denoising, on the other hand, folds in information of all lines from all exposures.  Here, it is sufficient to completely remove 10\% of exposures, a traditional training/validation fraction.  Since the information being removed varies between each test depending on its focus, we present the results per line, treating each line as an independent test.}

@@

\added{The per-line residuals from the denoising test also exhibit smaller spread than interpolation alone.  This suggests that the spectrograph truly is accurately represented by a low-dimensional model.  Recreating line positions using this model gives better line position estimates than treating each exposure independently.  The low-dimensional model does not incorporate noise from individual line measurements.  Returning more precise, denoised line positions results in smaller per-line residuals.}

@@

We tested \name -generated wavelengths on RV measurements using \expres\ observations of \target, which are presented in Table \ref{tab:rv}.  \target\ is a G0V star.  It is $4.8\, Gyr$ old and relatively quiet ($log R'_{HK} = -5.09$) \citep{brewer2016}.  We use 114 observations of \target\ taken between October 8, 2019 and March 5, 2020 with SNR 250.  Radial-velocity measurements were derived using a chunk-by-chunk, forward-modeling algorithm ran by the \expres\ team \citep{petersburg2020}.

Figure \ref{fig:rvs} compares the resultant RVs when using a classic, ninth-degree polynomial wavelength solution and an \name -generated wavelength model.  Using \name -generated wavelengths reduces the RMS of the entire data set from $1.17\, \mps$ with the classic wavelength solution to $1.05\, \mps$.  This is equivalent to removing an independent, additive noise component of $0.52\, \mps \,(=\sqrt{1.17^2-1.05^2})$.  \deleted{A reduction in RMS was observed universally in all data sets tested; other data sets showed a reduction in RMS between $0.2-0.5\, \mps$}.

\begin{deluxetable}{c c c}[t!]
\tabletypesize{\scriptsize}
\tablecaption{\expres\ RVs using \name\ Wavelengths \label{tab:rv}}
\tablehead{
\colhead{JD-2440000} & \colhead{RV $\mps$ } & \colhead{Error $\mps$ }
}
\startdata
 18764.4771  & 3.139  & 0.335 \\
 18764.4791  & 1.035  & 0.332 \\ 
 18764.4810  & 3.074  & 0.324 \\
 18771.4179  &-1.927  & 0.342 \\
          & \(\vdots\)  & \\
\enddata
\tablecomments{The full data set is available online}
\end{deluxetable}

@@

For the dimensionality reduction and denoising, the PCA could be replaced by a probabilistic PCA model or other probabilistic linear reduction methods, such as heteroscedastic matrix factorization (HMF)\citep{tsalmantza2012}.  It is also possible to move to non-linear reduction methods, like an auto-encoder or a normalizing flow \citep[e.g.][]{kramer1991,woodbridge2020}.  \added{Careful choice in the denoising model could enable \name\ to capture large-scale changes as well as small variations in calibration state.  This would bridge data broken up by significant format changes (e.g. replacing hardware, opening the spectrograph chamber, etc.).}

@@

The choice in interpolation method can help guide how many wavelength calibration images are truly needed.  It is unnecessary to take calibration images at times where the same information can be \added{reconstructed at the desired precision} by a well-chosen interpolation scheme.  For example, with the \expres\ data shown here, it is clear that nightly calibration images are needed, but for a linear trend, only two calibration images throughout the night are strictly required.

We also tested an implementation of \name\ where the $K$ principal components within a night were fit to a cubic \added{with respect to time rather than linearly interpolated}.  This emulates the current, polynomial-based wavelength solution implemented in the \expres\ pipeline, where polynomial fits to calibration files are interpolated to science exposure by fitting polynomial coefficients with respect to time to a cubic.  We found that \replaced{this method}{using a cubic in place of linear interpolation} returns comparable RV RMS for most targets, though appears to do better when a night has sparse calibration data.  \added{This suggests that the nightly behavior of \expres\ with respect to time is well described by a cubic function, but LFC exposures are typically taken with enough frequency that a linear interpolation provides a good approximation (see Figure \ref{fig:nightlyVariation}}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/intpx_tests.pdf}
\caption{Residuals from different interpolation schemes over pixels in echelle order 100.  ThAr lines, shown as blue circles, are used to construct a wavelength solution that is then evaluated at each LFC line, shown as blue vertical lines.  The residuals of each wavelength solution for a subset of the order is shown on the left.  Histograms of the residuals for each method for the complete order is shown on the right.  \deleted{Here, the y-axis is zoomed in to show the shape of each histogram more clearly.  }Note: there is a blended ThAr line at approximately pixel 5300, the right most ThAr lines plotted.}
\label{fig:xinterp}
\end{figure*}

@@

A simple linear interpolation would give erroneously low values everywhere.  Due to the dispersion intrinsic to echelle spectrographs, the wavelength change between pixels grows greater with greater wavelengths, \added{even within an order}.  This means that the function of wavelength vs. pixel across an echelle order will always be monotonically increasing and concave down everywhere.

\added{Though less of an issue with LFC lines,} a more classic cubic spline interpolation can run into issues with arc lines, which are irregularly spaced or even blended.  Close lines appearing at virtually the same pixel location but with different wavelengths could coerce a cubic spline into a very high slope.  \added{This is demonstrated by the green line in Figure \ref{fig:xinterp}, which shows the results of interpolating between ThAr lines rather than LFC lines.}  A line blend appears at approximately pixel 5300, causing the spline to twist to nearly vertical to account for both points.  This leads to huge deviations from the correct wavelengths around this line blend as the extraneously high slope of the spline is accounted for.

@@

As an empirical assessment of the needed line density, we removed LFC lines from the \expres\ data and calculated the per-line RMS of the returned wavelengths for the removed lines.  \added{These tests are similar to the interpolation test explained in Section \ref{sec:tests}, in that a fraction of lines are systematically removed from the analysis.  An increasing fraction of lines are removed to simulate different line densities.  For these line density tests, though, we are also implementing denoising unlike the pure interpolation test of Section \ref{sec:tests}.}

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.45\textwidth]{Figures/lfcDensityTest.pdf}
\caption{Per-line RMS as a function of spacing between lines in units of resolution element.  The average line spacing is calculated using the average distance between LFC lines across its wavelength range.}
\label{fig:density}
\end{figure}

@@

The calibration for any science exposure with a simultaneous reference can be determined by finding the amplitude of each basis vector that most closely recreates the calibration line positions through the reference fiber.  These amplitudes can then be used to recreate the calibration line positions through the science fiber as well.  This replaces the need to interpolate the basis vector amplitudes from calibration exposures to science exposures, something that is done with respect to time in the example implementation described in this paper.  The result is likely to be even more precise, as this method incorporates more data.  \added{This method, as with all analysis involving a simultaneous reference fiber, will work only as well as the reference fiber's ability to trace changes in the main science fiber.}

@@

\deleted{If the FSR varies very little, its value can in principle be captured by a PCA or similar method.  However, to interpolate this information from calibration exposures to science exposures would necessitate a simultaneous reference or interpolation with respect to some other housekeeping data that contains information on the value of the FSR.  If the FSR varies a lot, it becomes untenable to capture these changes with a PCA or similar method.  Instead, the resultant basis vectors of the calibration space would have to be conditioned on the value of the FSR.  This would require a non-trivial extension of \name.}

\added{Incorporating the FSR as part of the \name\ model will require introducing a free parameter to capture changes in the FSR independent of variation in an instrument's calibration state.  The calibration state can then be described with respect to mode number, which will be used to uniquely identify a calibration line across exposures rather than wavelength.  The FSR is then used to determine how the mode number of each line maps to wavelength for a given exposure.  The FSR must not vary so much that the change in this mode-number-to-wavelength mapping becomes non-linear.  This model would require a simultaneous reference or other housekeeping data that can be used to determine the FSR for every exposure.}

@@

For the results presented here, the data was broken up into different eras of stability based on where the principal component amplitudes showed huge deviations.  This was done visually, though there exists many change-point detection algorithms that could be used \citep{aminikhanghahi2017}.  There is a trade-off in including more exposures between introducing greater variation, but also more data to offer constraints that may be optimized.  Here, an era of stability was chosen  manually in order to focus on separating out time-domain intervals in which the data is relatively homogeneous, e.g. most exposures show the same calibration lines.  Homogeneity is, of course, implicitly required when implementing PCA.  \added{Different denoising models will be able to account for different amounts of stability or lack thereof.}

@@

\deleted{The endeavor to reach $0.1\,\mps$ radial-velocity precision necessitates improvements on all levels of spectroscopy, from the instrumentation to the data analysis.  Implementing \name, a hierarchical, non-parametric model for wavelength calibration, is a step towards reducing the error contribution from data reduction pipelines.}

@@

Similarly, PCA is just one of many possible dimensionality-reduction methods.  We chose to implement \name\ using PCA here for simplicity and computational tractability.  If \name\ was updated to a full probabilistic model, the PCA along with the interpolation model would have to be upgraded to something with better probabilistic properties.  \added{Other denoising methods may be more robust to large changes, allowing all calibration images ever taken with an instrument to be used to construct the accessible calibration state regardless of hardware adjustments.}  Further discussion of other implementations of \name\ can be found in Section \ref{sec:choices}.

@@

Once we have defined a calibration space that captures all possible degrees of freedom for a stabilized spectrograph, there are many options for pinpointing where the spectrograph is located within that calibration space.  Good housekeeping data, such as temperature or pressure could be used in addition to or instead of time (as mentioned in Section \ref{sec:choice_avt}).  Telemetry that is seen to be correlated with the calibration state of the spectrograph can even be added to the data used to construct the low-dimensional basis.  Furthermore, all exposures taken with the spectrograph in principle contains information about the calibration state of the spectrograph.  Theoretically, tellurics, lines in science exposures, or just the trace positions themselves could also be used to determine an instrument's calibration state, \replaced{negating the need for any additional wavelength-calibration source beyond defining the calibration space.}{thereby providing free simultaneous calibration information.}

>>>>

\begin{abstract}\noindent%
\Name\ is a non-parametric, hierarchical framework for precision wavelength-calibration of spectrographs.  It is designed with the needs of extreme-precision radial velocity (\eprv) in mind, which require that instruments be calibrated or stabilized to better than $10^{-4}$ pixels.  Instruments vary along only a few dominant degrees of freedom, especially \eprv\ instruments that feature highly stabilized optical systems and detectors.  \Name\ takes advantage of this property by using all calibration data to construct a low-dimensional representation of all accessible calibration states for an instrument.  \Name\ also takes advantage of laser frequency combs or etalons, which generate a dense set of stable calibration points.  This density permits the use of a non-parametric wavelength solution that can adapt to any instrument or detector oddities better than parametric models, such as a polynomial.  We demonstrate the success of this method with data from the \textsl{EXtreme PREcision Spectrograph} (\expres), which uses a laser frequency comb.  When wavelengths are assigned to laser comb lines using \name, the RMS of the residuals is about five times lower than wavelengths assigned using polynomial fits to individual exposures.  Radial-velocity measurements of HD~34411 showed a reduction in RMS scatter over a 10-month time baseline from $1.17$ to $1.05\, \mps$.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- instrumentation: detectors -- techniques: spectroscopic -- techniques: radial velocities -- methods: data analysis -- methods: statistical}
%\section*{~}\pagebreak
\section{Introduction} 
Precise, radial-velocity programs have been fruitful in finding and characterizing extra-solar planets \citep[e.g.][]{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $50,000-100,000$, which correspond to line widths on the order of $3000\,\mps$.  The state of the art RV precision reached $1\, \mps$ by 2016 \citep{fischer2016}.  The newest generation of instruments aim  to reach $0.1\,\mps$ precision, the required precision to detect terrestrial worlds.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).  Two next-generation spectrographs, \expres\ and \espresso, have been commissioned for more than a year and are demonstrating $<0.1 \mps$ instrumental errors and $\sim 0.2\, \mps$ errors on stars \citep{pepe2013, jurgenson2016, blackman2020, petersburg2020, brewer2020, mascareno2020}.

@@

The second simple idea follows from the observation that most physical systems have only a few dominant degrees of freedom, meaning most spectrographs vary along only a small number of axes in ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  This is particularly true of \eprv\ instruments, which are equipped with stringent environmental stabilizing.  The thermomechanical stability of these instruments reduces the variations they experience to something that can be represented by a low-dimensional framework.  That is, spectrographs, especially stabilized ones, should have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the calibration space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the small, accessible part of calibration space the spectrograph was located for each exposure.

@@

The challenge is to learn these basis functions, $g_k(\lambda,m)$, from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern the basis functions.  In this paper, we present one implementation of \name\ using principal component analysis (PCA) \citep{pearson1901, jolliffe2016}.  A PCA is justifiable in the limit where exposures have very high signal-to-nose ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

@@

\Name\ assumes that line positions have been identified ``correctly,'' as in the position of a calibration line is determined the same way as the position of a stellar line when extracting RVs.  This also inherently assumes that the calibration lines are not subject to an effect that the science exposures are not, for example differences in charge transfer inefficiency, non-linearities in the PSF, etc.  We caution that systemic errors or large uncertainties in fitting line positions easily propagate through to biases in the wavelength models returned by \name.  For more discussion, see Section \ref{sec:discussion}.

@@

\begin{algorithm}
\SetAlgoLined
\KwData{line positions $x(\lambda,m,n)$ for each exposure $n$, with wavelengths $\lambda$ and echelle orders $m$}
\KwResult{Basis vectors of the low-dimensional calibration space $g_k(\lambda,m)$ and location of exposures in calibration space expressed by amplitudes $a_{n,k}$}
\While{change in missing or outlier line centers $>$ 0.01\%}
{
	$g_0(\lambda,m) = \overline{x(\lambda,m,n)}$\;
	using Singular-Value Decomposition, find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x(\lambda,m,n)-g_0(\lambda,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(\lambda,m) = V$\;
	$x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)$ for $x(\lambda,m,n)$ = \code{NaN} where $K$ is a a small integer
	}
\caption{Dimensionality Reduction and De-Noising}
\end{algorithm}

@@

In the implementation of \name\ presented here, the amplitudes of the principal components are interpolated linearly with respect to time.  This is discussed more in Section \ref{sec:choice_avt}.  Let a prime denote values related to a science exposure $n'$ for which we want wavelengths.  We use linearly interpolated magnitudes, $a_{n',k}$ at time $t_{n'}$ to construct the calibration state of the spectrograph for that point in time.  Using interpolated amplitudes, $a_{n',k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, a new set of calibration lines, $x'(\lambda,m,n')$ can be constructed for any exposure as formulated in Equation \ref{eq:excl_wsol}.

@@

After experiments, we found that a cubic-spline interpolation that enforces monotonicity, such as a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a parameterized function, while the enforced monotonicity allows the wavelength solution, $\lambda(x',m,n')$, to be invertible and prevents spurious deviations that may befall a cubic spline.  Choices in interpolation scheme, $K$, and other tests are further discussed in Section \ref{sec:choice_wvp}.

\begin{algorithm}
\SetAlgoLined
\KwData{the fiducial calibration of the spectrograph $g_0(\lambda,m)$; magnitudes of the principal components for each exposure $a_{n,k}$; basis vectors spanning the calibration space of the spectrograph $g_k(\lambda,m)$; }
\KwResult{Wavelengths for detector positions $x'(m,n')$ of exposure $n'$ with time $t_{n'}$, where the prime is used to denote values relevant to this new exposure}

@@

The implementation of \name\ described here is hosted on GitHub\footnote{\url{https://www.github.com/lilyling27/excalibur}}.

@@

In order to comfortably satisfy the assumption that there exists only low-order variation, which is needed for \name, we used exposures from after the LFC stabilized following servicing in summer 2019, where the photonic crystal fiber was replaced and the polarization was changed to shift the wavelength range of the LFC redwards.  In the results presented here, we use 1227 LFC exposures and 78 ThAr exposures taken between October 14 and December 18, 2019 on 29 unique nights.

@@

The polynomial and interpolation tests remove the same 50\% of lines from each exposure while the denoising test completely removes a randomly selected 10\% of calibration exposures and their associated line position measurements.  Errors from interpolation will be localized, extending only to neighboring lines.  We therefore aggressively remove every other line to ensure we are capturing these local effects.  The PCA denoising, on the other hand, folds in information of all lines from all exposures.  Here, it is sufficient to completely remove 10\% of exposures, a traditional training/validation fraction.  Since the information being removed varies between each test depending on its focus, we present the results per line, treating each line as an independent test.

@@

\begin{figure}[b]
\centering
\includegraphics[width=.45\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and theoretical wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The per line RMS as defined in Equation \ref{eq:rms} is given in the top-right corner in each method's corresponding color.  Incorporating denoising returns the smallest spread in residuals.}
\label{fig:testHists}
\end{figure}

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/lineResids2D.pdf}
\caption{Residuals of a single LFC exposure plotted with respect to detector position (as defined by echelle order and x-pixel) for parametric (top) and non-parametric (bottom) wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the theoretical wavelength for each line, given in units of $\mps$.  High-order structure, i.e. vertical stripes and patchiness, is apparent in the residuals to a polynomial wavelength solution, which assumes smoothness.}
\label{fig:resid2d}
\end{figure*}

The per-line residuals from the denoising test also exhibit smaller spread than interpolation alone.  This suggests that the spectrograph truly is accurately represented by a low-dimensional model.  Recreating line positions using this model gives better line position estimates than treating each exposure independently.  The low-dimensional model does not incorporate noise from individual line measurements.  Returning more precise, denoised line positions results in smaller per-line residuals.

@@

We tested \name -generated wavelengths on RV measurements using \expres\ observations of \target, which are presented in Table \ref{tab:rv}.  \target\ is a G0V star.  It is $4.8\, Gyr$ old and relatively quiet ($log R'_{HK} = -5.09$) \citep{brewer2016}.  Because \target\ has no known planets and should have a smaller contribution from stellar signals, it is a good star to test the effects of the wavelength calibration on the RMS of the returned RVs.  We use 114 observations of \target\ taken between October 8, 2019 and March 5, 2020 with SNR 250.  Radial-velocity measurements were derived using a chunk-by-chunk, forward-modeling algorithm ran by the \expres\ team \citep{petersburg2020}.

Figure \ref{fig:rvs} compares the resultant RVs when using a classic, ninth-degree polynomial wavelength solution and an \name -generated wavelength model.  Using \name -generated wavelengths reduces the RMS of the entire data set from $1.17\, \mps$ with the classic wavelength solution to $1.05\, \mps$.  This is equivalent to removing an independent, additive noise component of $0.52\, \mps \,(=\sqrt{1.17^2-1.05^2})$.

We conducted a direct test of a classically-generated wavelength solution with \name -wavelengths on four other data sets.  All targets showed a reduction in or comparable RV RMS.  The results from these data sets can not be interpreted as directly as with \target, though, due to larger contributions from stellar variability, known planets, etc.  As completely mitigating these different effects is out of scope for this paper, we focus here on the results with \target.

\begin{deluxetable}{c c c}[t!]
\tabletypesize{\scriptsize}
\tablecaption{\expres\ RVs using \name\ Wavelengths \label{tab:rv}}
\tablehead{
\colhead{JD-2440000} & \colhead{RV $\mps$ } & \colhead{Error $\mps$ }
}
\startdata
 18764.4771  & 3.139  & 0.335 \\
 18764.4791  & 1.035  & 0.332 \\ 
 18764.4810  & 3.074  & 0.324 \\
 18771.4179  &-1.927  & 0.342 \\
 18771.4196 & 2.688 &	0.357 \\
          & \(\vdots\)  & \\
\enddata
\tablecomments{The full data set is available online}
\end{deluxetable}

@@

For the dimensionality reduction and denoising, the PCA could be replaced by a probabilistic PCA model or other probabilistic linear reduction methods, such as heteroscedastic matrix factorization (HMF)\citep{tsalmantza2012}.  It is also possible to move to non-linear reduction methods, like a Guassian process latent-variable model, an auto-encoder, a normalizing flow \citep[e.g.][]{kramer1991,woodbridge2020}.  Using a non-linear denoising model could enable \name\ to capture large-scale changes as well as small variations in calibration state.

@@

The choice in interpolation method can help guide how many wavelength calibration images are truly needed.  It is unnecessary to take calibration images at times where the same information can be reconstructed at the desired precision by a well-chosen interpolation scheme.  For example, with the \expres\ data shown here, it is clear that nightly calibration images are needed, but for a linear trend, only two calibration images throughout the night are strictly required.

We also tested an implementation of \name\ where the $K$ principal components within a night were fit to a cubic with respect to time rather than linearly interpolated.  This emulates the current, polynomial-based wavelength solution implemented in the \expres\ pipeline, where polynomial fits to calibration files are interpolated to science exposure by fitting polynomial coefficients with respect to time to a cubic.  We found that using a cubic in place of linear interpolation returns comparable RV RMS for most targets, though appears to do better when a night has sparse calibration data.  This suggests that the nightly behavior of \expres\ with respect to time is well described by a cubic function, but LFC exposures are typically taken with enough frequency that a linear interpolation provides a good approximation (see Figure \ref{fig:nightlyVariation})

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/intpx_tests.pdf}
\caption{Residuals from different interpolation schemes over pixels in echelle order 100.  ThAr lines, shown as blue circles, are used to construct a wavelength solution that is then evaluated at each LFC line, shown as blue vertical lines.  The residuals of each wavelength solution for a subset of the order is shown on the left.  Histograms of the residuals for each method for the complete order is shown on the right.  Note: there is a blended ThAr line at approximately pixel 5300, the right most ThAr lines plotted.}
\label{fig:xinterp}
\end{figure*}

@@

A simple linear interpolation would give erroneously low values everywhere.  Due to the dispersion intrinsic to echelle spectrographs, the wavelength change between pixels grows greater with greater wavelengths, even within an order.  This means that the function of wavelength vs. pixel across an echelle order will always be monotonically increasing and concave down everywhere.

Though less of an issue with LFC lines, a more classic cubic spline interpolation can run into issues with arc lines, which are irregularly spaced or even blended.  Close lines appearing at virtually the same pixel location but with different wavelengths could coerce a cubic spline into a very high slope.  This is demonstrated by the green line in Figure \ref{fig:xinterp}, which shows the results of interpolating between ThAr lines rather than LFC lines.  A line blend appears at approximately pixel 5300, causing the spline to twist to nearly vertical to account for both points.  This leads to huge deviations from the correct wavelengths around this line blend as the extraneously high slope of the spline is accounted for.

@@

As an empirical assessment of the needed line density, we removed LFC lines from the \expres\ data and calculated the per-line RMS of the returned wavelengths for the removed lines.  These tests are similar to the interpolation test explained in Section \ref{sec:tests}, in that a fraction of lines are systematically removed from the analysis.  An increasing fraction of lines are removed to simulate different line densities.  For these line density tests, though, we are also implementing denoising unlike the pure interpolation test of Section \ref{sec:tests}.

\begin{figure}[b]
\centering
\includegraphics[width=0.45\textwidth]{Figures/lfcDensityTest.pdf}
\caption{Per-line RMS as a function of spacing between lines in units of resolution element.  The average line spacing is calculated using the average distance between LFC lines across its wavelength range.}
\label{fig:density}
\end{figure}

@@

@@

The calibration for any science exposure with a simultaneous reference can be determined by finding the amplitude of each basis vector that most closely recreates the calibration line positions through the reference fiber.  These amplitudes can then be used to recreate the calibration line positions through the science fiber as well.  This replaces the need to interpolate the basis vector amplitudes from calibration exposures to science exposures, something that is done with respect to time in the example implementation described in this paper.  The result is likely to be even more precise, as this method incorporates more data.  This method, as with all analysis involving a simultaneous reference fiber, will work only as well as the reference fiber's ability to trace changes in the main science fiber.

@@

Incorporating the FSR as part of the \name\ model will require introducing a free parameter to capture changes in the FSR independent of variation in an instrument's calibration state.  The calibration state can then be described with respect to mode number, which will be used to uniquely identify a calibration line across exposures rather than wavelength.  The FSR is then used to determine how the mode number of each line maps to wavelength for a given exposure.  The FSR must not vary so much that the change in this mode-number-to-wavelength mapping becomes non-linear.  This model would require a simultaneous reference or other housekeeping data that can be used to determine the FSR for every exposure.

@@

For the results presented here, the data was broken up into different eras of stability based on where the principal component amplitudes showed huge deviations.  This was done visually, though there exists many change-point detection algorithms that could be used \citep{aminikhanghahi2017}.  There is a trade-off in including more exposures between introducing greater variation, but also more data to offer constraints that may be optimized.  Here, an era of stability was chosen  manually in order to focus on separating out time-domain intervals in which the data is relatively homogeneous, e.g. most exposures show the same calibration lines.  Homogeneity is, of course, implicitly required when implementing PCA.  Different denoising models will be able to account for different amounts of stability or lack thereof.

@@

@@

Similarly, PCA is just one of many possible dimensionality-reduction methods.  We chose to implement \name\ using PCA here for simplicity and computational tractability.  PCA is a good option because the instrument changes here are small enough that a linear model is an approrpiate representation of the changes.  If \name\ was updated to a full probabilistic model, the PCA along with the interpolation model would have to be upgraded to something with better probabilistic properties.  Other, nonlinear denoising methods may be more robust to large changes, allowing all calibration images ever taken with an instrument to be used to construct the accessible calibration state regardless of hardware adjustments.  Further discussion of other implementations of \name\ can be found in Section \ref{sec:choices}.

@@

Once we have defined a calibration space that captures all possible degrees of freedom for a stabilized spectrograph, there are many options for pinpointing where the spectrograph is located within that calibration space.  Good housekeeping data, such as temperature or pressure could be used in addition to or instead of time (as mentioned in Section \ref{sec:choice_avt}).  Telemetry that is seen to be correlated with the calibration state of the spectrograph can even be added to the data used to construct the low-dimensional basis.  Furthermore, all exposures taken with the spectrograph in principle contains information about the calibration state of the spectrograph.  Theoretically, tellurics, lines in science exposures, or just the trace positions themselves could also be used to determine an instrument's calibration state, thereby providing free simultaneous calibration information.
